---
title: "01_cleandata"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Install libraries
```{r cars}
library(mctoolsr)
library(plyr) # always load before tidyverse to avoid conflicts with dplyr
library(tidyverse) # lazily load all of tidyverse, just in case I want to use it.
library(vegan)
#library(plotly)
library(readr)
library("RColorBrewer")
library(viridis)
library(wesanderson)
```

## Set up input and output directories
```{r Set up input and output directories, include = FALSE}
project.fp <- "/Users/coin8046/Desktop/FiererLabNotebook_copy/SunflowerGxE/R.analyses_ITS"
raw_data.fp <- file.path(project.fp, "00_raw.data")
raw_metadata.fp <- "/Users/coin8046/Desktop/FiererLabNotebook_copy/SunflowerGxE/r_analyses/00_raw.data"
clean_data.fp <- file.path(project.fp, "01_clean.data")
    figures_01_clean.fp <- file.path(clean_data.fp, "figures")
    outputs_01_clean.fp <- file.path(clean_data.fp, "outputs")

if (!dir.exists(clean_data.fp)) {dir.create(clean_data.fp, recursive = TRUE)}    
if (!dir.exists(figures_01_clean.fp)) {dir.create(figures_01_clean.fp, recursive = TRUE)}
if (!dir.exists(outputs_01_clean.fp)) {dir.create(outputs_01_clean.fp, recursive = TRUE)}
```

## Read in data
```{r Read in data, include = FALSE}
# ASV generated from DADA2, with sequence inference on pooled samples
tax_table.R1_fp <- paste0(raw_data.fp, '/seqtab_wTax_mctoolsr.R1.txt')

# mapping file with (minimal) sample info
map_fp <- paste0(raw_metadata.fp, '/sunflower_metadata1.txt')
#tmp_map <- read_table(file = paste0(raw_data.fp, '/mapping_file_mm_ITS.txt'), col_names = TRUE)

# load data via mctoolsr
input.R1 <- load_taxa_table(tax_table.R1_fp, map_fp)


## extract just map file for later plotting
map.r <- input.R1$map_loaded %>% rownames_to_column(var = 'sample_id')
```

### read in other metadata
```{r}
# read in other metadata
## sunflower species
spc.entries <- read.csv(paste0(raw_metadata.fp, '/sunflower_entries.csv'))
## sclerotinia resistance
scl.res <- read.csv(paste0(raw_metadata.fp, '/SclIncidenceMeansCarrington2017.csv'))
## gps coordinates
gps.loc <- read.csv(paste0(raw_metadata.fp, '/microbiomeGxEsites_gpscoords.csv'))
```

### Adjust mapping file
```{r}
# adjust mapping file to add condition, soil type, sterility columns
map.rr <- map.r %>% 
  left_join(data.frame(ReadCount.raw = colSums(input.R1$data_loaded)) %>%
               rownames_to_column(), by = c("sample_id" = "rowname")) %>% 
  # left_join(data.frame(ReadCount.filt = colSums(input_filt.f$data_loaded)) %>%
  #              rownames_to_column(), by = c("sample_id" = "rowname")) %>% 
  mutate(Rownames = sample_id) %>% 
  column_to_rownames(var = "Rownames")

# merge species id info 
spc.entries.rr <- spc.entries %>% 
  mutate(plot_id = paste0(Site, "_", Plot))

# fix entry ID 
scl.res1 <- scl.res %>% 
  mutate(Entry = gsub(Line.name, pattern = "(.{2})(.*)", replacement = "\\1 \\2"))


map.rrr <- map.rr %>% 
  mutate(plot_id = paste0(Location, "_", plot)) %>% 
  left_join(spc.entries.rr, by = "plot_id") %>% 
  left_join(gps.loc, by = "Location") %>% 
  left_join(scl.res1, by = "Entry")

input.R1$map_loaded <- map.rrr %>% 
  column_to_rownames(var = "sample_id")

# input.R1$map_loaded <- map
```

### Filtering
#### R1
```{r}
# calculate raw reads per sample / total
reads0 <- input.R1$data_loaded %>% colSums() %>% sum()
# 19980040

### remove taxa not classified at phylum level
input.f1 <- filter_taxa_from_input(input = input.R1, taxa_to_remove = "NA", at_spec_level = 2)
# 11592 taxa removed

# calculate reads after first filer
reads1 <- input.f1$data_loaded %>% colSums() %>% sum()
# 15796644

# count taxa not identified at phylum level
input.NA <- filter_taxa_from_input(input = input.R1, taxa_to_keep = "NA", at_spec_level = 2)
reads2 <- input.NA$data_loaded %>% colSums() %>% sum()
# 4183396


reads2/reads0
# 0.2093788
reads1/reads0
# 0.7906212


### remove low read count ASVs (< 15 reads across dataset)
ASV_reads <- input.f1$data_loaded %>% rowSums()
low_asvs.df <- ASV_reads %>% 
  as.data.frame() %>% 
  filter(. < 15) %>% 
  rownames_to_column(var = "ASV")
low_asvs <- unlist(low_asvs.df$ASV)

input.f2 <- filter_taxa_from_input(input = input.f1, taxa_IDs_to_remove = low_asvs )

### remove really low read count samples
sort(colSums(input.f2$data_loaded))
input.f <- filter_samples_by_counts(input = input.f2, min_seqs = 600)
```


### What are the read depth statistics?
```{r}
### R1 only

cat('Summary read depth statistics: raw reads, R1 only')
raw.R1_reads_per_sample.p <- sort(colSums(input.R1$data_loaded))
summary(raw.R1_reads_per_sample.p )
   # Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   #    1   22799   31632   32488   43552   81817 

sum(raw.R1_reads_per_sample.p)
# 19980040
nrow(input.R1$data_loaded)
# 25223

cat('Summary read depth statistics: filtered data, R1 only')
filt.R1_reads_per_sample.p <- sort(colSums(input.f$data_loaded))
summary(filt.R1_reads_per_sample.p )
   # Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   # 1025   18978   25505   26802   34434   62012 

sum(filt.R1_reads_per_sample.p)
# 15759805
nrow(input.f$data_loaded)
# 7081
```

### Normalization
#### Relative abundance
```{r}
### convert filtered dataframe to relative abundance

input.f.rl <- convert_to_relative_abundances(input.f)
```

#### Rarefaction
```{r}
sort(colSums(input.f$data_loaded))

input_rar1 <- single_rarefy(input.f, 4000)
# 587 samples remaining
input_rar2 <- single_rarefy(input.f, 10000)
# 562 samples remaining



# input_rar1 <- single_rarefy(input_R1.f, 1000)
# input_rar2 <- single_rarefy(input_R1.f, 10000)

```


##### plotting settings
```{r}
pair.pal2 <- c("#FFFF99", "#1F78B4", "#33A02C", "#E31A1C", "#FF7F00", "#6A3D9A", "#A6CEE3","#B2DF8A","#FB9A99","#FDBF6F","#CAB2D6")
```

### Blank check
```{r}
input_blank <- filter_data(input = input.R1, filter_cat = "Sample_type", keep_vals = "blank")

blank_df <- input_blank$data_loaded %>% 
  mutate(asv_reads = rowSums(.[1:ncol(input_blank$data_loaded)])) %>% 
  mutate(asv_ubiq = rowSums(.[1:ncol(input_blank$data_loaded-1)] !=0)) %>% 
  rownames_to_column(var = "ASV") %>% 
  inner_join(input_blank$taxonomy_loaded, by = c("ASV" = "taxonomy7")) %>% 
  mutate(ASV2 = ASV) %>% 
  select(-c("taxonomy1", "taxonomy8", "taxonomy4")) %>% 
  column_to_rownames(var = "ASV")
  
nrow(blank_df)
# 55

input_check <- filter_samples_by_counts(input = input.R1, min_seqs = 100)
input_check.rl <- convert_to_relative_abundances(input = input_check)
```




### Sample summary info
```{r}
### calculating sample summary info
input.f_samples = filter_data(input.f, 'Sample_type', keep_vals = c("sample"))


#num reads / sample
summary(colSums(input.f_samples$data_loaded))
  # Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  #  4397   18984   25518   26846   34450   62012


#num asvs / sample
summary(colSums(input.f_samples$data_loaded !=0))
   # Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   # 70.0   154.0   188.0   186.6   216.0   327.0 

```

#### colors for plotting
```{r}
rainbow15 <- wes_palette("Zissou1", 15, type = "continuous")
rainbow10 <- wes_palette("Zissou1", 10, type = "continuous")
pal11 <- c("#D53E4F",  "#FDAE61", "#ABDDA4","#9E0142", "#66C2A5", "#3288BD", "#5E4FA2", "#F46D43","#E6F598",  "#5BA9BC", "#708090")
#smkmtn6 <- park_palette("SmokyMountains", 6)
```


### Compare Samples and Blanks
#### NMDS Visualization
```{r, include=FALSE}
#### Visualization: NMDS relative abundance (all > 100 reads)
## distance matrix for plotting
sb_transformed <- t(sqrt(input_check.rl$data_loaded))
sb.dm <- vegdist(sb_transformed, method = "bray", na.rm = TRUE)
sb.nmds <- metaMDS(sb.dm, k = 2, trymax = 50)

## plot nmds
nmds_all <- plot_ordination(input = input_check.rl, ordination_axes = sb.nmds$points, color_cat = 'Sample_type')+
  ggtitle("NMDS plot of all samples - R1 (relative abundance)")
nmds_all

#### Visualization: NMDS relative abundance (by location)
## distance matrix for plotting
sall_transformed <- t(sqrt(input_rar2$data_loaded))
dm <- vegdist(sall_transformed, method = "bray", na.rm = TRUE)
sall.nmds <- metaMDS(dm, k = 2, trymax = 50)

input_rar2$map_loaded$Rep <- as.factor(input_rar2$map_loaded$Rep)
## plot nmds
nmds_all <- plot_ordination(input = input_rar2, ordination_axes = sall.nmds$points, color_cat = 'Location', shape_cat = 'Rep')+
  ggtitle("NMDS plot of all samples (ITS reads - R1 only, rarefied 10k), colored by location")
nmds_all

#### Visualization: NMDS relative abundance (by sclerotinia resistance)
## plot nmds
nmds_scl <- plot_ordination(input = input_rar2, ordination_axes = sall.nmds$points, color_cat = 'Average.of.percent.incidence')+
  scale_color_viridis(option = "magma", begin = 0.2, end = 0.8, discrete = FALSE)+
  ggtitle("NMDS plot of all samples (ITS reads - R1 , rarefied 10k), colored by Sclerotinia resistance")
nmds_scl
```

#### individual locations
```{r}
input_Grandin <- filter_data(input = input_rar2, filter_cat = "Location", keep_vals = "Grandin")
input_Lindsborg.dry <- filter_data(input = input_rar2, filter_cat = "Location", keep_vals = "Lindsborg-dryland")
input_Lindsborg.irr <- filter_data(input = input_rar2, filter_cat = "Location", keep_vals = "Lindsborg-irrigated")
input_McLaughlin <- filter_data(input = input_rar2, filter_cat = "Location", keep_vals = "McLaughlin")
input_Land.inst <- filter_data(input = input_rar2, filter_cat = "Location", keep_vals = "Land Institute")
```
##### NMDS visualizations
```{r}
## NMDS ordinations by location
location_input <- input_Land.inst
location_input$map_loaded$Rep <- as.factor(location_input$map_loaded$Rep)

## distance matrix for plotting
sb_transformed.loc <- t(sqrt(location_input$data_loaded))
dm.loc <- vegdist(sb_transformed.loc, method = "bray")
sb.loc.nmds <- metaMDS(dm.loc, k = 2, trymax = 1000)

#names(input_filt_rar$map_loaded)

## plot nmds
nmds_loc <- plot_ordination(input = location_input, ordination_axes = sb.loc.nmds$points, color_cat = 'Entry', shape_cat = "Rep", hulls = TRUE)+ #Entry, Location, Average.of.percent.incidence, LONG, LAT
  ggtitle("NMDS plot of Grandin - colored by genotype")+
  scale_color_manual(values = pal11)+
  scale_fill_manual(values = pal11)
  #scale_color_viridis(option = "magma", begin = 0.2, end = 0.8, discrete = TRUE)
  #theme(legend.position = "none")
nmds_loc
```



### Richness and top taxa
```{r}
plot_diversity(input = input.R1_slrsol, variable = 'sample_type_2', metric = 'richness')+
  scale_fill_manual(values=pair.pal2)+
  geom_point()

plot_diversity(input = input.R1_rhizo, variable = 'condition', metric = 'richness')+
  scale_fill_manual(values=pair.pal2)+
  ggtitle("rhizosphere richness by condition")

plot_diversity(input = input.R1_seed, variable = 'condition', metric = 'richness')+
  scale_fill_manual(values=pair.pal2)+
  ggtitle("seed richness by condition")
```

### Save cleaned / filtered dataframes
```{r}
## clean raw data
saveRDS(object = input_R1.f, file = paste0(outputs_01_clean.fp, "/input_filt.RDS"))
input_R1.f <- readRDS(file = paste0(outputs_01_clean.fp, "/input_filt.RDS"))

## relative abundance
saveRDS(object = input.f.rl, file = paste0(outputs_01_clean.fp, "/input_filt_rl.RDS"))

## clean data rarefied (1000,10000 reads)
saveRDS(object = input.R1_rar1, file = paste0(outputs_01_clean.fp, "/input_rar1k.RDS"))
saveRDS(object = input.R1_rar2, file = paste0(outputs_01_clean.fp, "/input_rar10k.RDS"))
```