---
title: "05_climate"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries
```{r}
library(mctoolsr)
library(plyr) # always load before tidyverse to avoid conflicts with dplyr
library(dplyr)
library(tidyverse) # lazily load all of tidyverse, just in case I want to use it.
library(vegan)
library(MASS)
library(wesanderson)
library(viridis)
library(edgeR)
library(car)
library(performance)
library(lme4)
library(nlme)
library(lmerTest)
library(compositions)
library(ggplot2)
library(ggrepel)

library(sp)
library(raster)
library(sf)
library(tmap)
#library(maptools)
#library(rgdal)
library(dismo)
library(usmap)
library(ggspatial)
library(spData)
library(ggthemes)
library(patchwork)
library(ggforce)
library(ggThemeAssist)
library(devtools)
library(readxl)

set.seed(10)
```

### Set up input and output directories
```{r Set up input and output directories, include = FALSE}
project.fp <- "/Users/coin8046/Desktop/FiererLabNotebook_copy/SunflowerGxE/r_analyses"
raw_data.fp <- file.path(project.fp, "00_raw.data")
clean_data.fp <- file.path(project.fp, "01_clean.data")
    figures_01_clean.fp <- file.path(clean_data.fp, "figures")
    outputs_01_clean.fp <- file.path(clean_data.fp, "outputs")

climate_05.fp <- file.path(project.fp, "05_ClimateData")
    figures_05.fp <- file.path(climate_05.fp, "figures")
    outputs_05.fp <- file.path(climate_05.fp, "outputs")

 if (!dir.exists(climate_05.fp)) {dir.create(climate_05.fp, recursive = TRUE)}
 if (!dir.exists(figures_05.fp)) {dir.create(figures_05.fp, recursive = TRUE)}
 if (!dir.exists(outputs_05.fp)) {dir.create(outputs_05.fp, recursive = TRUE)}
```

### get climate data
## from https://chelsa-climate.org/exchelsa-extended-bioclim/
```{r}
#downloaded from chelsa-climate.org Nov 2023
myRasterCMI <- raster(paste0(raw_data.fp, "/CHELSA_cmi_mean_1981-2010_V.2.1.tif"))

myRasterGGD <- raster(paste0(raw_data.fp, "/CHELSA_gdd0_1981-2010_V.2.1.tif"))
```

### Read in 16S data
```{r}
## cleaned data
input_filt <- readRDS(paste0(outputs_01_clean.fp, "/input_filt.RDS"))

## rarefied data
input_rar <- readRDS(paste0(outputs_01_clean.fp, "/input_filt_rar9k.RDS"))

## relative abundance data
input_relab <- readRDS(paste0(outputs_01_clean.fp, "/input_filt_relab.RDS"))

```

### filter to decrease super rare ASVS
```{r}
#remove ASVs present in fewer than 30 samples
low.tmp <- input_filt$data_loaded %>% 
  mutate(ubiq = rowSums(. !=0)) %>% 
  filter(ubiq < 30)

lowASVs <- row.names(low.tmp)

input_filt.rl <- filter_taxa_from_input(input = input_relab, taxa_IDs_to_remove =  lowASVs)
nrow(input_filt.rl$data_loaded)
# < 30:
# 16968 taxa removed
# 10960 remaining

input_filt2 <- filter_taxa_from_input(input = input_filt, taxa_IDs_to_remove =  lowASVs)
# < 30:
# 16968 taxa removed
# 10960 remaining
```

#### get metadata temporary dataframe
```{r}
map.r <- input_filt2$map_loaded

map.r$State.x %>% unique()
names(map.r)
```

### Setup data for map
NOTE: this code is adapted from Kyle Keepers
```{r}
#Get a shapefile using "raster" package, then create a vector of states, then
#subset by that vector
myShapeFile <- raster::getData("GADM",country="USA",level=1)
myStates <- c("Texas","Kansas","Colorado","North Dakota","South Dakota", "Oklahoma", "Nebraska", "Minnesota", "New Mexico", "Wyoming")
state.sub <- myShapeFile[as.character(myShapeFile@data$NAME_1) %in% myStates, ]

## get CMI data in order
#Crop and mask. Not 100% sure what these steps do, but necessary
precip.sub <- raster::crop(myRasterCMI,extent(state.sub))
precip.sub <- raster::mask(precip.sub,state.sub)

## get GGD (growing degree days) data in order
#Crop and mask. Not 100% sure what these steps do, but necessary
ggd.sub <- raster::crop(myRasterGGD,extent(state.sub))
ggd.sub <- raster::mask(ggd.sub,state.sub)

#Create vectors of the coordinates
#then convert them to spatial points (??) so we can plot them on the map
myCoords <- dplyr::select(map.r,LONG,LAT,Site) %>%
  na.omit() %>%
  unique() %>%
  as.data.frame()

#Do some converting
myWestSpatialPoints <- SpatialPointsDataFrame(coords = myCoords[,c(1,2)], data = myCoords, proj4string = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))
```

### Plot map
Note:this code was adatped from Kyle Keepers
```{r}
set.seed(Sys.time())
map.cmi <- tm_shape(precip.sub,alpha=0.3)+
  tm_raster(title = "Mean CMI",n=6)+
  tm_layout(legend.position = c("left","bottom"),legend.title.size = 2,legend.text.size = 1.25, legend.width = 2.5)+
  tm_shape(state.sub)+
  tm_borders(lwd=2)+
  tm_shape(myWestSpatialPoints)+
  tm_text("Site",col="black",size=1.25, auto.placement = TRUE)+
  tm_symbols(shape=21,size=0.8,col="#F8766D",border.col = "black",border.lwd = 2)

map.ggd <- tm_shape(ggd.sub,alpha=0.3)+
  tm_raster(title = "Growing Degree Days > 0˚C",n=6)+
  tm_layout(legend.position = c("left","bottom"),legend.title.size = 2,legend.text.size = 1.25, legend.width = 2.5)+
  tm_shape(state.sub)+
  tm_borders(lwd=2)+
  tm_shape(myWestSpatialPoints)+
  tm_text("Site",col="black",size=1.25, auto.placement = TRUE)+
  tm_symbols(shape=21,size=0.8,col="#F8766D",border.col = "black",border.lwd = 2)
 
```


### Extract Climate Variables (Climate Moisture Index, GDD0) values for locations
NOTE: this code was adapted from Kyle Keepers
```{r}
#First, subset location lat/long cols and remove any NAs
coords <- map.r %>% 
  dplyr::select(LONG,LAT,Location) %>%
  na.omit() %>%
  unique() %>%
  as.data.frame() %>% 
  remove_rownames() %>% 
  column_to_rownames("Location")


#Project the lat/long coordinates onto discrete cells in the raster (I think)
points.cmi <- SpatialPoints(coords, proj4string = myRasterCMI@crs)
points.gdd <- SpatialPoints(coords, proj4string = myRasterGGD@crs)

#Extract the climate moisture index values at each of the 12 prairies
CMI <- raster::extract(myRasterCMI, points.cmi)
GDD <- raster::extract(myRasterGGD, points.gdd)

#Add the CMI values to the "coords" array
bind_coords_cmi <- cbind(coords, CMI) %>%
  cbind(GDD) %>% 
  rownames_to_column("Location")

#Add them to the master data array for use throughout the rest of this script
map.rr <- map.r %>%
  rownames_to_column(var = "sampleID") %>% 
  left_join(bind_coords_cmi) %>% 
  mutate(cmi.scl = scale(CMI)) %>% 
  mutate(gdd.scl = scale(GDD)) %>% 
  mutate(Entry = as.factor(Entry)) %>% 
  mutate(Location = as.factor(Location)) %>% 
  column_to_rownames(var = "sampleID")
```

### add metadata to input (fir later downstream analyses)
```{r}
input_filt2$map_loaded <- map.rr

input_filt$map_loaded <- map.rr

saveRDS(object = input_filt2, file = paste0(outputs_05.fp, "/input.filt2C.RDS"))
saveRDS(object = input_filt, file = paste0(outputs_05.fp, "/input.filtC.RDS"))
  
```




